Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-20 output/latex/manuscript.tex
latex: remove unnecessary references in main text.
fix letter symbol
rename Authors list section
move tables and figure captions at the end

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=1in,left=0.5in,right=0.5in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

\hypertarget{authors}{%
\subsection{Authors}\label{authors}}

@@

  \textbf{Casey S. Greene}\textsuperscript{16,2} \\
  \textsuperscript{\protect\hyperlink{correspondence}{✉}}

@@

\leavevmode\vadjust pre{\hypertarget{correspondence}{}}%
✉ --- Correspondence possible via GitHub Issues
or email to
Casey S. Greene \textless casey.s.greene@cuanschutz.edu\textgreater.

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.pdf}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top), with 53 independent samples expressed in Neutrophils, 59 in Granulocytes, and 20 in Whole blood, 56 in PBMC, 8 in mDCs, 29 in Monocytes, and 5 in Epithelial cells (the boxplot shows the 25th, 50th and 75th percentiles while the whiskers extend to the minimum/maximum values).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:lv246}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/lvs_analysis/lv246/lv246.svg}
\caption{\textbf{Tissues and traits associated with a gene module related to lipid metabolism (LV246).}
\textbf{a)} Top cell types/tissues in which LV246's genes are expressed.
Values in the \(y\)-axis come from matrix \(\mathbf{B}\) in the MultiPLIER models (Figure \ref{fig:entire_process}b, see Methods).
In the \(x\)-axis, cell types/tissues are sorted by the maximum sample value.
\textbf{b)} Gene-trait associations (unadjusted \(p\)-values from S-MultiXcan\textsuperscript{\protect\hyperlink{ref-1FFzCXo1s}{24}}; threshold at -log(\(p\))=10) and colocalization probability (fastENLOC) for the top traits in LV246.
The top 40 genes in LV246 are shown, sorted by their LV weight (matrix \(\mathbf{Z}\)), from largest (the top gene \emph{SCD}) to smallest (\emph{FAR2});
\emph{DGAT2} and \emph{ACACA}, in boldface, are two of the six high-confidence genes in the lipids-increasing gene set from the CRISPR screen.
Cardiovascular-related traits are in boldface.
SGBS: Simpson Golabi Behmel Syndrome;
CH2DB: CH2 groups to double bonds ratio;
HDL: high-density lipoprotein;
RCP: locus regional colocalization probability.
}\label{fig:lv246}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:drug_disease:roc_pr}{%
\centering
\includesvg[width=0.8\textwidth,height=\textheight]{images/drug_disease_prediction/roc_pr_curves.svg}
\caption{\textbf{Drug-disease prediction performance for gene-based and LV-based approaches.}
The receiver operating characteristic (ROC) (left) and the precision-recall curves (right) for a gene-based and LV-based approach.
``Random'' refers to the average precision of a hundred classifiers with randomly permuted scores, where the error band represents the 95\% confidence interval.
AUC: area under the curve; AP: average precision.}\label{fig:drug_disease:roc_pr}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:lv116:cell_types}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/lvs_analysis/lv116/lv116-cell_types.svg}
\caption{\textbf{Top cell types/tissues where LV116's genes are expressed.}
Values in the \(y\)-axis come from matrix \(\mathbf{B}\) in the MultiPLIER models (Figure \ref{fig:entire_process}b).
In the \(x\)-axis, cell types/tissues are sorted by the maximum sample value.
The figure shows a clear immune response with cell types under different stimuli.
MS: multiple sclerosis;
HSV: treated with herpes simplex virus;
WNV: infected with West Nile virus;
IFNa: treated with interferon-alpha;
HMDM: human peripheral blood mononuclear cell-derived macrophages;
Salm: infected with \emph{Salmonella typhimurium};
Yers: infected with \emph{Yersinia pseudotuberculosis};
ISM: Interferon Signature Metric;
SLE: Systemic lupus erythematosus.}\label{fig:lv116:cell_types}
}
\end{figure}

@@

\begin{table}
\caption{LVs among the top 20 contributors to the prediction of niacin for five cardiovascular diseases. ``Heart attack, angina, stroke or hypertension'' refers to the UK Biobank data-field 6150. GWAS sample size: Atherosclerosis (361,194 in total and 566 cases), Chronic ischaemic heart disease (361,194 in total and 12,769 cases), Heart attack, angina, stroke or hypertension (360,420 in total and 253,565 cases), Ischaemic heart disease/wide definition (361,194 in total and 20,857 cases), High cholesterol/self-reported (361,141 in total and 43,957 cases). \label{tbl:niacin:cardio:top_lvs}}\label{tbl:niacin:cardio:top_lvs}%\tabularnewline
\centering
\begin{tabular}{lll}
\toprule
LV & Cell type & Disease \\
\midrule
LV116 & Immune cells, skin & Atherosclerosis (ICD10 I70) \\
& & Chronic ischaemic heart disease (ICD10 I25) \\
& & Heart attack, angina, stroke or hypertension \\
& & Ischaemic heart disease (wide definition) \\
LV931 & Immune cells & Atherosclerosis (ICD10 I70) \\
& & Heart attack, angina, stroke or hypertension \\
& & Ischaemic heart disease (wide definition) \\
LV246 & Adipose tissue, liver & Atherosclerosis (ICD10 I70) \\
& & High cholesterol (self-reported) \\
& & Ischaemic heart disease (wide definition) \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\hypertarget{fig:clustering:design}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_design.svg}
\caption{\textbf{Cluster analysis on traits using the latent gene expression representation.}
\textbf{a)} The projection of TWAS results on 3,752 traits into the latent gene expression representation is the input data to the clustering process.
A linear (PCA) and non-linear (UMAP) dimensionality reduction techniques were applied to the input data, and five different clustering algorithms processed all data versions.
These algorithms derive partitions from the data using different parameters (such as the number of clusters), leading to an ensemble of 4,428 partitions.
Then, a distance matrix is derived by counting how many times a pair of traits was grouped in different clusters across the ensemble.
Finally, a consensus function is applied to the distance matrix to generate consolidated partitions with different numbers of clusters (from 2 to \(\sqrt{n}\approx\) 60).
These final solutions were represented in the clustering tree (Figure \ref{fig:clustering:tree}).
\textbf{b)} The clusters found by the consensus function were used as labels to train a decision tree classifier on the original input data, which detects the LVs that better differentiate groups of traits.}\label{fig:clustering:design}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:clustering:tree}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_tree.svg}
\caption{\textbf{Clustering tree using multiple resolutions for clusters of traits.}
Each row represents a partition/grouping of the traits, and each circle is a cluster from that partition.
The number of clusters goes from 5 to 29.
Arrows indicate how traits in one cluster move across clusters from different partitions.
Most of the clusters are preserved across different resolutions, showing highly stable solutions even with independent runs of the clustering algorithm.
RDW: red cell (erythrocyte) distribution width;
BMI: body mass index;
WC: waist circumference;
HC: hip circumference;
RA: rheumatoid arthritis;
SLE: systemic lupus erythematosus;
HTN: Hypertension;
IBD: inflammatory bowel disease;
SCZ: Schizophrenia;
CAD: Coronary artery disease;
AD: Alzheimer's disease;
The full lists of traits in each cluster in the last five partitions of the tree (from \(k=16\) to \(k29\)) are in Supplementary Files 3-7.}\label{fig:clustering:tree}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:clustering:heatmap}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/clustering/global_clustermap-plain.svg}
\caption{\textbf{Cluster-specific and general transcriptional processes associated with different diseases.}
The plot shows a submatrix of \(\hat{\mathbf{M}}\) for the main trait clusters at \(k\)=29, considering only LVs (rows) that are well-aligned with at least one pathway.}\label{fig:clustering:heatmap}
}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-SiobXsoB}{}}%
\CSLLeftMargin{117. }%
\CSLRightInline{Barbeira, A. N. \emph{et al.} \href{https://doi.org/10.1002/gepi.22346}{Fine‐mapping and QTL tissue‐sharing information improves the reliability of causal gene identification}. \emph{Genetic Epidemiology} \textbf{44}, 854--867 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-GPHGnFRN}{}}%
\CSLLeftMargin{118. }%
\CSLRightInline{\emph{\href{https://github.com/hakyimlab/summary-gwas-imputation}{Harmonization and Imputation Overview}}. (hakyimlab, 2022).}

\leavevmode\vadjust pre{\hypertarget{ref-VKYS05n1}{}}%
\CSLLeftMargin{119. }%
\CSLRightInline{Homo sapiens (ID 232177) - BioProject - NCBI. \url{https://www.ncbi.nlm.nih.gov/bioproject/PRJNA232177}.}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=1in,left=0.5in,right=0.5in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\usepackage[misc]{ifsym}

@@

\hypertarget{authors}{%
\subsection{Author list}\label{authors}}

@@

  \textbf{Casey S. Greene}\textsuperscript{16,2}
  \textsuperscript{\protect\hyperlink{correspondence}{\Letter}} \\

@@

\leavevmode\vadjust pre{\hypertarget{correspondence}{}}%
{\Letter} --- Correspondence possible via GitHub Issues
or email to
Casey S. Greene \textless casey.s.greene@cuanschutz.edu\textgreater.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\hypertarget{tables}{%
\subsection{Tables}\label{tables}}

\begin{table}[h]
\caption{LVs among the top 20 contributors to the prediction of niacin for five cardiovascular diseases. ``Heart attack, angina, stroke or hypertension'' refers to the UK Biobank data-field 6150. GWAS sample size: Atherosclerosis (361,194 in total and 566 cases), Chronic ischaemic heart disease (361,194 in total and 12,769 cases), Heart attack, angina, stroke or hypertension (360,420 in total and 253,565 cases), Ischaemic heart disease/wide definition (361,194 in total and 20,857 cases), High cholesterol/self-reported (361,141 in total and 43,957 cases). \label{tbl:niacin:cardio:top_lvs}}\label{tbl:niacin:cardio:top_lvs}%\tabularnewline
\centering
\begin{tabular}{lll}
\toprule
LV & Cell type & Disease \\
\midrule
LV116 & Immune cells, skin & Atherosclerosis (ICD10 I70) \\
& & Chronic ischaemic heart disease (ICD10 I25) \\
& & Heart attack, angina, stroke or hypertension \\
& & Ischaemic heart disease (wide definition) \\
LV931 & Immune cells & Atherosclerosis (ICD10 I70) \\
& & Heart attack, angina, stroke or hypertension \\
& & Ischaemic heart disease (wide definition) \\
LV246 & Adipose tissue, liver & Atherosclerosis (ICD10 I70) \\
& & High cholesterol (self-reported) \\
& & Ischaemic heart disease (wide definition) \\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{figurecaptions}{%
\subsection{Figure Legends/Captions}\label{figurecaptions}}

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
% \includegraphics[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.pdf}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top), with 53 independent samples expressed in Neutrophils, 59 in Granulocytes, and 20 in Whole blood, 56 in PBMC, 8 in mDCs, 29 in Monocytes, and 5 in Epithelial cells (the boxplot shows the 25th, 50th and 75th percentiles while the whiskers extend to the minimum/maximum values).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

\begin{figure}
\hypertarget{fig:lv246}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/lvs_analysis/lv246/lv246.svg}
\caption{\textbf{Tissues and traits associated with a gene module related to lipid metabolism (LV246).}
\textbf{a)} Top cell types/tissues in which LV246's genes are expressed.
Values in the \(y\)-axis come from matrix \(\mathbf{B}\) in the MultiPLIER models (Figure \ref{fig:entire_process}b, see Methods).
In the \(x\)-axis, cell types/tissues are sorted by the maximum sample value.
\textbf{b)} Gene-trait associations (unadjusted \(p\)-values from S-MultiXcan\textsuperscript{\protect\hyperlink{ref-1FFzCXo1s}{24}}; threshold at -log(\(p\))=10) and colocalization probability (fastENLOC) for the top traits in LV246.
The top 40 genes in LV246 are shown, sorted by their LV weight (matrix \(\mathbf{Z}\)), from largest (the top gene \emph{SCD}) to smallest (\emph{FAR2});
\emph{DGAT2} and \emph{ACACA}, in boldface, are two of the six high-confidence genes in the lipids-increasing gene set from the CRISPR screen.
Cardiovascular-related traits are in boldface.
SGBS: Simpson Golabi Behmel Syndrome;
CH2DB: CH2 groups to double bonds ratio;
HDL: high-density lipoprotein;
RCP: locus regional colocalization probability.
}\label{fig:lv246}
}
\end{figure}

\begin{figure}
\hypertarget{fig:drug_disease:roc_pr}{%
\centering
% \includesvg[width=0.8\textwidth,height=\textheight]{images/drug_disease_prediction/roc_pr_curves.svg}
\caption{\textbf{Drug-disease prediction performance for gene-based and LV-based approaches.}
The receiver operating characteristic (ROC) (left) and the precision-recall curves (right) for a gene-based and LV-based approach.
``Random'' refers to the average precision of a hundred classifiers with randomly permuted scores, where the error band represents the 95\% confidence interval.
AUC: area under the curve; AP: average precision.}\label{fig:drug_disease:roc_pr}
}
\end{figure}

\begin{figure}
\hypertarget{fig:lv116:cell_types}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/lvs_analysis/lv116/lv116-cell_types.svg}
\caption{\textbf{Top cell types/tissues where LV116's genes are expressed.}
Values in the \(y\)-axis come from matrix \(\mathbf{B}\) in the MultiPLIER models (Figure \ref{fig:entire_process}b).
In the \(x\)-axis, cell types/tissues are sorted by the maximum sample value.
The figure shows a clear immune response with cell types under different stimuli.
MS: multiple sclerosis;
HSV: treated with herpes simplex virus;
WNV: infected with West Nile virus;
IFNa: treated with interferon-alpha;
HMDM: human peripheral blood mononuclear cell-derived macrophages;
Salm: infected with \emph{Salmonella typhimurium};
Yers: infected with \emph{Yersinia pseudotuberculosis};
ISM: Interferon Signature Metric;
SLE: Systemic lupus erythematosus.}\label{fig:lv116:cell_types}
}
\end{figure}

\begin{figure}
\hypertarget{fig:clustering:design}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_design.svg}
\caption{\textbf{Cluster analysis on traits using the latent gene expression representation.}
\textbf{a)} The projection of TWAS results on 3,752 traits into the latent gene expression representation is the input data to the clustering process.
A linear (PCA) and non-linear (UMAP) dimensionality reduction techniques were applied to the input data, and five different clustering algorithms processed all data versions.
These algorithms derive partitions from the data using different parameters (such as the number of clusters), leading to an ensemble of 4,428 partitions.
Then, a distance matrix is derived by counting how many times a pair of traits was grouped in different clusters across the ensemble.
Finally, a consensus function is applied to the distance matrix to generate consolidated partitions with different numbers of clusters (from 2 to \(\sqrt{n}\approx\) 60).
These final solutions were represented in the clustering tree (Figure \ref{fig:clustering:tree}).
\textbf{b)} The clusters found by the consensus function were used as labels to train a decision tree classifier on the original input data, which detects the LVs that better differentiate groups of traits.}\label{fig:clustering:design}
}
\end{figure}

\begin{figure}
\hypertarget{fig:clustering:tree}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/clustering/clustering_tree.svg}
\caption{\textbf{Clustering tree using multiple resolutions for clusters of traits.}
Each row represents a partition/grouping of the traits, and each circle is a cluster from that partition.
The number of clusters goes from 5 to 29.
Arrows indicate how traits in one cluster move across clusters from different partitions.
Most of the clusters are preserved across different resolutions, showing highly stable solutions even with independent runs of the clustering algorithm.
RDW: red cell (erythrocyte) distribution width;
BMI: body mass index;
WC: waist circumference;
HC: hip circumference;
RA: rheumatoid arthritis;
SLE: systemic lupus erythematosus;
HTN: Hypertension;
IBD: inflammatory bowel disease;
SCZ: Schizophrenia;
CAD: Coronary artery disease;
AD: Alzheimer's disease;
The full lists of traits in each cluster in the last five partitions of the tree (from \(k=16\) to \(k29\)) are in Supplementary Files 3-7.}\label{fig:clustering:tree}
}
\end{figure}

\begin{figure}
\hypertarget{fig:clustering:heatmap}{%
\centering
% \includesvg[width=1\textwidth,height=\textheight]{images/clustering/global_clustermap-plain.svg}
\caption{\textbf{Cluster-specific and general transcriptional processes associated with different diseases.}
The plot shows a submatrix of \(\hat{\mathbf{M}}\) for the main trait clusters at \(k\)=29, considering only LVs (rows) that are well-aligned with at least one pathway.}\label{fig:clustering:heatmap}
}
\end{figure}
